Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-03 demystify.tex
Updates from Overleaf
updates

\paragraph{Why CS depts do not waive application fee?}  Note that some universities do waive application fee, however most do not.  In any case, application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, if the application process were free for everyone, most good CS programs would receive an overwhelming number of applications to review.

@@

Most PhD programs will require at least \textbf{two LORs}. Having a strong letter from an internationally recognized researcher will greatly strengthen your application. In fact, a strong letter from a well-known researcher can outweigh the lack of publications.

@@

\paragraph{How to help your LOR writer} Your LoR writer can ask you to provide additional information to help them write your letter.  For example, they might ask your GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and can complement that with their own perspective.  It might also be a good idea to provide them such information even when they do not ask, especially if you have not interacted with them much or have not done much research with them.

@@

There are various guides on writing SOP, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidences. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..).

You should also talk about things that adcom members might not know about and can help make you different than the other 1000 applicants, e.g.,  your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

This is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part, so do not skip it.

Finally, you should get your SOP reviewed by LoR writers, professors (especially those who have served in adcom), or even graduate PhD students (because they have been through this process).

@@

Graduating from top universities \emph{that we recognize} helps. For example, if you are an international student and your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and  likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

What can you do about this? You can explain about your school in your SOP (and also ask your LOR writer to do that). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\paragraph{Timeline} The timeline for interviews and notification letters varies.  Faculty setup interviews based on their (busy) schedule. Do not be surprised if you get an interview invitation in the last minute. Some schools do not do interviews at all.

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will have to make your decision by around \emph{April 15} (see \href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why?}).

@@

\begin{table}
  \centering
  \small
  \caption{Different types of PhD funding. All cover tuition, insurance, and stipend.}\label{tab:funding}
  \begin{tabular}{c|c|c|c}
    \toprule
    &\textbf{TA}&\textbf{RA}&\textbf{Fellowship}\\
    \midrule
    \textbf{From} & School & Profs. & School/External\\
    \textbf{For}                  & Teaching Assist.       & Research                        & Research                              \\
    \textbf{Cover All?} & Yes                      & Yes                             & Yes                                   \\
    \textbf{Summer?}              & No                       & Maybe                           & Yes                                   \\
    \midrule
    \textbf{Pros}                 & Research Freedom         & Get to do research              & Research Freedom                      \\
    \textbf{Cons}                 & Teaching Duties           & Research Restriction & Competitive, Limited             \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\textbf{How to get Fellowship?}   Many schools provide fellowships to attract admitted students. You likely will not need to do anything and the adcom will recommend such fellowships to admitted students. Some schools automatically offer a fellowship to all accepted students, while others only award it to a limited number of admitted students, such as the top percentile.

You likely need to apply for external fellowships including those from the US government and tech companies (e.g., Google, Microsoft, Facebook).  Fellowships from US government would likely require US citizenship or residency. % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan, so it might be a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

Here are some general advice to find professor.  Assuming you're not familiar with any particular one, then first, search for profs. that share similar research interests. You can learn about faculty and their level of research activity through the faculty's website or even from CSRankings.  For example, in CSRankings, if you want to work with PL, you can search for those publish in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you do research about that professor and contact them (\autoref{sec:contact}).

@@

\section{Why did I get rejected?}
Many possible reasons!  Other than having weak profile (e.g., research potentials, GPA, LoRs, SOP), here are some that you might not expect:
\begin{itemize}
  \item  \emph{Your English scores (e.g., IELTS or TOEFL) are not high enough}.  While professors might not care much about these, there is a certain minimum set by the college or the university that you need to pass to get admitted, especially for TA funding (e.g., you need to be able to communicate well with students as a TA).  Moreover, while professors are willing to fight to get you admitted, they might be reluctant to go against the standard requirement of the university or college on English scores.  Note this is mostly for international students, as domestic one do not have to take these tests.

  \item \emph{No faculty available to advise you}.  You could have an excellent profile, but if there is no faculty who is willing to advise you (e.g., they already have too many students), you will not be admitted. This is actually for your own good because you do not want to be in a program where you cannot find an adviser.
\end{itemize}

>>>>

\paragraph{Why we do not waive application fee?}  This is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, if the application process were free for everyone, we would receive an overwhelming number of applications to review.

@@

Most PhD programs will require at least \textbf{two LORs}. Having a strong letter from an internationally recognized researcher will greatly strengthen your application. In fact, a strong letter from a well-known researcher can outweight the lack of publications.

@@

@@

There are various guides on writing SOP, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. Also, talk about things that adcom members might not know about, e.g., your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants (e.g., see \autoref{sec:improve-your-chance} for increasing your admission chance).

% \begin{commentbox}[Hung:]
%   I think asking for a “research vision” from a Ph.D. applicant is too much. Even people graduated with a Ph.D. have a great difficulty in explicating their research vision. In my opinion, you should focus on showing (not telling) your research passion (if you do have).
% \end{commentbox}

Finally, this is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part, so do not skip it.

@@

Graduating from top universities \emph{that we recognize} helps. For example, if your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and  likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

What can you do about this? You can explain about your school in your statement (ask your LOR writer to do that too). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS PhD programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\paragraph{Timeline} The timeline for interviews and notification letters varies.  Faculty setup interviews based on their (very busy) schedule. Do not be surprised if you get an interview invitation in the last minute. Again, some schools do not do interviews at all.

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will have to make your decision by around \emph{April 15}.

@@

\begin{table}
  \centering
  \footnotesize
  \caption{Different types of PhD funding. All Covered: includes tuition, insurance, and stipend.}\label{tab:funding}
  \begin{tabular}{c|c|c|c}
    \toprule
    &\textbf{TA}&\textbf{RA}&\textbf{Fellowship}\\
    \midrule
    \textbf{From} & School & Profs. & School/External\\
    \textbf{For}                  & Teaching Assist.       & Research                        & Research                              \\
    \textbf{Cover All?} & Yes                      & Yes                             & Yes                                   \\
    \textbf{Summer?}              & No                       & Maybe                           & Yes                                   \\
    \midrule
    \textbf{Pros}                 & Research Freedom         & Get to do research              & Research Freedom                      \\
    \textbf{Cons}                 & Teaching Duties           & Research Restriction & Competitive, Limited             \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\textbf{How to get Fellowship?} You need to apply for them.  Many schools provide such fellowships to attract admitted students.
External fellowships including from the US government and tech companies (e.g., Google, Microsoft, Facebook).  Fellowships from US government would likely require US citizenship or residency. % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.

Prestigious external fellowships typically require a clear and good research plan, so it is a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

Here are some general advice to find professor.  First, search for profs. that share similar research interests. You can learn about faculty and their level of research activity through the faculty's website or even from CSRankings.  For example, in CSRankings, if you want to work with PL, you can search for those publish in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.   You can also go to their website, look at their papers and projects.  In any case, you should contact the professor and talk to them. For example, you can (cold) email them and ask about their research and if they are taking new students (\autoref{sec:contact}), and schedule appointments to chat with them (office hrs are often the best time).

@@

\section{Why did I get rejected?}
Many possible reasons!  Other than not having very strong profile (e.g., research potentials, GPA, LoRs, SOP), here are some that you might not expect:
\begin{itemize}

\item \emph{You are not a good fit to the program}.  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if there is no faculty who is willing to advise you (e.g., they are on sabbatical, NSF leave, or already have too many students currently), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

For example, in the 2024 admission cycle, there is a huge surge in students who are interested in AI and NLP (thanks, ChatGPT!). However, our AI/NLP faculty cannot take more students and therefore we cannot admit many applicants, many of whom have excellent profiles (e.g., papers in top conferences, strong LoRs, etc.).

\item  \emph{Your English scores (e.g., IELTS or TOEFL) are not high enough}.  While professors might not care much about these, there is a certain minimum set by the college or the university that you need to pass to get admitted, especially for TA funding (e.g., you need to be able to communicate well with students as a TA).  Thus, while professors are willing to argue for your case, they might be reluctant to go against the standard requirement of the university on English scores.  Note this is mostly for international students, as domestic ones (or those from these \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{countries}) do not have to take these tests.

\end{itemize}
